---
title: "Template"
author: "Studentnames and studentnumbers here"
date: "`r Sys.Date()`"
output: pdf_document
---

# Set-up your environment

```{r package_install, include=FALSE}
install.packages("tidyverse")
install.packages("yaml")
install.packages("rmarkdown")
install.packages("WDI")
install.packages("eurostat")
install.packages("ggrepel")
```

```{r packages}
require(tidyverse)
require(rmarkdown)
require(yaml)
require(WDI)
require(eurostat)
require(ggrepel)
```

# Title Page

LE Vi Trang Raghoebar Monisha Isabel Lust Ermias Gebremichael Pritish Dewan Pritish Dewan Noel Tripathi Caelin Rohiemon

Tutorial 6 Group 2

Simon Loop

# Part 1 - Identify a Social Problem

Use APA referencing throughout your document. [Here's a link to some explanation.](https://www.mendeley.com/guides/apa-citation-guide/)

## 1.1 Describe the Social Problem

Include the following:

- Why is this relevant?

- ...

# Part 2 - Data Sourcing

## 2.1 Load in the data

```{r loading_data and retrieve only data of European countries}

options(timeout = 600)
# Eurostat 
ai_raw <- get_eurostat("isoc_eb_ai", time_format = "num")

# Filter out EU countries, fix the 2 misaligned EU codes and drop aggregates
eu_codes <- ai_raw %>%
  distinct(geo) %>%
  mutate(iso2c = recode(geo, "EL" = "GR", "UK" = "GB")) %>%
  filter(nchar(iso2c) == 2, iso2c != "EA") %>%
  pull(iso2c)

# World Bank: download ONCE, then reuse the saved copy
if (file.exists("wb_raw.rds")) {
  wb_raw <- readRDS("wb_raw.rds")
} else {
  wb_raw <- WDI(
    country   = eu_codes,
    indicator = c(total_unemp = "SL.UEM.TOTL.ZS",
                  youth_unemp = "SL.UEM.1524.ZS"),
    start = 1991, end = 2024, extra = TRUE
  )
  saveRDS(wb_raw, "wb_raw.rds")
}
```

## 2.2 Provide a short summary of the dataset(s)

```{r}
head(ai_raw) 
str(ai_raw)
summary(ai_raw)
names(ai_raw)

head(wb_raw)
str(wb_raw)
summary(wb_raw)
names(wb_raw)

```

## 2.3 Describe the type of variables included

Think of things like:

- Do the variables contain health information or SES information?

- Have they been measured by interviewing individuals or is the data coming from administrative sources?

The AI data comes from Eurostats that shows business wealth and resources . The unemployment data comes from World Bankdata that shows the financial trouble and living standards .the two main variables are  Total unemployment rate and youth unemploymentent. the youth unemployment is used as the sub-population to see how young workers are affected by AI data usage in businesses.
Where: European countries only.
When: Years 2021 to 2025 ( but 2022 not included). This timeline captures the exact period where the post-pandemic labor market met the massive generative AI boom.
 
*For the sake of this example, I will continue with the assignment...*

# Part 3 - Quantifying

## 3.1 Data cleaning

Say we want to include only larger distances (above 2) in our dataset, we can filter for this.

```{r data_cleaning}

wb_clean <- wb_raw %>%
  filter(year == 2021 | year == 2022 | year == 2023 | year == 2024)

ai_clean <- ai_raw %>%
  filter(indic_is == "E_AI_TANY")
  
colnames(ai_clean)
colnames(wb_clean)

ai_clean <- ai_clean %>%
  rename(
    year = 'TIME_PERIOD',
    iso2c = 'geo'
  )


merged <- inner_join(ai_clean, wb_clean, by = c("iso2c", "year"))
relationship = "many-to-many"



```

Please use a separate 'R block' of code for each type of cleaning. So, e.g. one for missing values, a new one for removing unnecessary variables etc.

## 3.2 Generate necessary variables

Variable 1

```{r gen_var1}

wb_cleaned <- wb_cleaned %>%
  mutate(youth_ratio = youth_unemp / total_unemp)

```

Variable 2

```{r gen_var2}

combined <- wb_cleaned %>%
  left_join(ai_clean, by = c("iso2c" = "geo", "year" = "TIME_PERIOD")) %>%
mutate(ai_group = case_when(
  values < 10 ~ "Low",
  values < 30 ~ "Medium",
  TRUE ~ "High"
))
names(wb_cleaned)

```


Table with new variables

```{r}
combined %>%
  group_by(ai_group) %>%
  summarise(
    n = n(),
    mean_total_unemp = mean(total_unemp, na.rm = TRUE),
    mean_youth_ratio = mean(youth_ratio, na.rm = TRUE)
  )
```



## 3.3 Visualize temporal variation

```{r}
ggplot(wb_cleaned, aes(x= year)) + 
         geom_line(aes(y=mean_total), color = "red") +
          geom_line(aes(y=mean_youth), color = "blue")
```
ai_group - bins the continuous values into Low, Medium, and High, so we can compare youth unemployment outcomes across different levels of AI adotion.

## 3.4 Visualize spatial variation

```{r visualise_map}

ggplot (merged, aes(x=ai_intensity, 
                    y= youth_unemployment,
                    size= total_unemployment,
                    label=country)) +
  geom_point(alpha=0.7, color = "stealblue") + 
  geom_text_repel(size=3, max.overlaps= 20)+
  scale_size_continuous(name="total unemployment (%)") +
  
  labs(x="AI intensity", 
       y="youth unemployment(%)", 
       title= "AI intensity and unemployment in EU countries") + theme_minimal()
                 
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.5 Visualize sub-population variation

```{r}
combined %>%
  filter(year == 2024, !is.na(ai_group)) %>%
  mutate(ai_group = factor(ai_group, levels = c("Low", "Medium", "High"))) %>%
  ggplot(aes(x = ai_group, y = youth_unemp)) +
  geom_boxplot(na.rm = TRUE, fill = "antiquewhite", color = "navy", outlier.color = "aquamarine", varwidth = TRUE) +
  labs(title = "Youth unemployment by level of AI adoption (2024)",
       x = "AI adoption group", y = "Youth unemployment (%)") +
  theme_minimal()
```



## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

Use APA referencing throughout your document.
