---
title: "Template"
author: "Studentnames and studentnumbers here"
date: "`r Sys.Date()`"
output: pdf_document
---

# Set-up your environment

```{r package_install, include=FALSE}
install.packages("tidyverse")
install.packages("yaml")
install.packages("rmarkdown")
install.packages("WDI")
```

```{r packages}
require(tidyverse)
require(rmarkdown)
require(yaml)
require(WDI)
```

# Title Page

LE Vi Trang
Raghoebar Monisha 
Isabel Lust
Ermias Gebremichael
Pritish Dewan 

Tutorial 6 Group 2

Simon Loop

# Part 1 - Identify a Social Problem

Use APA referencing throughout your document. [Here's a link to some explanation.](https://www.mendeley.com/guides/apa-citation-guide/)

## 1.1 Describe the Social Problem

Include the following:

-   Why is this relevant?

-   ...

# Part 2 - Data Sourcing

## 2.1 Load in the data

Preferably from a URL, but if not, make sure to download the data and store it in a shared location that you can load the data in from. Do not store the data in a folder you include in the Github repository!

```{r loading_data}

wb_raw <- WDI(
  country   = "all",
  indicator = c(total_unemp = "SL.UEM.TOTL.ZS",
                youth_unemp = "SL.UEM.1524.ZS"),
  start     = 1991,            # WB modelled-ILO series effectively starts here
  end       = 2024,
  extra     = TRUE
)

# Keep individual countries only (drop WLD, regions, income groups).
# If you want the World total for a context line, keep wb_raw instead.
wb_countries <- wb_raw %>%
  filter(region != "Aggregates") %>%
  select(country, iso3c = iso3c, region, income, year,
         total_unemp, youth_unemp) %>%
  arrange(country, year)

# ============================================================
# 2. EUROSTAT  (eurostat package hits ec.europa.eu directly)
# ============================================================
# time_format = "num" gives a numeric year column.
ai_raw <- get_eurostat("isoc_eb_ai", time_format = "num")

# --- INSPECT the dimensions before filtering -------------------------------
# Eurostat dimension CODES vary by dataset and change occasionally, so we
# DISCOVER them at runtime rather than hard-coding values that may be wrong.
dim_cols <- setdiff(names(ai_raw), c("values", "TIME_PERIOD", "time"))
cat("\nEurostat isoc_eb_ai dimensions and their codes:\n")
print(lapply(ai_raw[dim_cols], function(x) sort(unique(as.character(x)))))

# Human-readable labels (handy for confirming which code means what)
ai_labelled <- label_eurostat(ai_raw)

# --- TEMPLATE filter --- VERIFY the codes against the printout above --------
# Typical target: % of enterprises with 10+ employees using >=1 AI technology.
# Common codes are below, but CONFIRM them in the inspection output and edit:
#   size class column is often 'sizen_r2' with "GE10" = 10+ employees
#   unit column is often 'unit' with "PC_ENT" = percentage of enterprises
#   indicator column is often 'indic_is' (pick the "at least one AI tech" code)
ai_clean <- ai_raw
if ("sizen_r2" %in% names(ai_raw)) ai_clean <- filter(ai_clean, sizen_r2 == "GE10")
if ("unit"     %in% names(ai_raw)) ai_clean <- filter(ai_clean, unit == "PC_ENT")
# After confirming the indicator code, uncomment and set it, e.g.:
# ai_clean <- filter(ai_clean, indic_is == "E_AI_TTAI")

ai_clean <- ai_clean %>%
  rename(year = TIME_PERIOD, ai_use_pct = values) %>%
  arrange(geo, year)
```


## 2.2 Provide a short summary of the dataset(s)

```{r}
head(dataset)
```

In this case we see 28 variables, but we miss some information on what units they are in. We also don't know anything about the year/moment in which this data has been captured.

``` r
inline_code = TRUE
```

These are things that are usually included in the metadata of the dataset. For your project, you need to provide us with the information from your metadata that we need to understand your dataset of choice.

## 2.3 Describe the type of variables included

Think of things like:

-   Do the variables contain health information or SES information?

-   Have they been measured by interviewing individuals or is the data coming from administrative sources?

*For the sake of this example, I will continue with the assignment...*

# Part 3 - Quantifying

## 3.1 Data cleaning

Say we want to include only larger distances (above 2) in our dataset, we can filter for this.

```{r data_cleaning}
mean(dataset$percollege)
```

Please use a separate 'R block' of code for each type of cleaning. So, e.g. one for missing values, a new one for removing unnecessary variables etc.

## 3.2 Generate necessary variables

Variable 1

```{r gen_var1}

```

Variable 2

```{r gen_var2}

```

## 3.3 Visualize temporal variation

```{r}

```

## 3.4 Visualize spatial variation

```{r visualise_map}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.5 Visualize sub-population variation

What is the poverty rate by state?

```{r visualise_subpopulations}
dataset$inmetro <- dataset$inmetro %>% as.factor()
# Boxplot of poverty rate by state using the 'midwest' dataset
ggplot(dataset, aes(x = inmetro, y = percadultpoverty)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Poverty Rates by Metropolitan status (Midwest counties)",
    x = "Metropolitan Area",
    y = "Poverty Rate of Adults (%)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right"
  )
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

Use APA referencing throughout your document.
