---
title: "Template"
author: "Studentnames and studentnumbers here"
date: "`r Sys.Date()`"
output: pdf_document
---

# Set-up your environment

```{r package_install, include=FALSE}
install.packages("tidyverse")
install.packages("yaml")
install.packages("rmarkdown")
install.packages("WDI")
install.packages("eurostat")
install.packages("ggrepel")
install.packages("tinytex")
```

```{r packages}
require(tidyverse)
require(rmarkdown)
require(yaml)
require(WDI)
require(eurostat)
require(ggrepel)
```

# Title Page

LE Vi Trang 
Raghoebar Monisha 
Isabel Lust 
Ermias Gebremichael 
Pritish Dewan 
Noel Tripathi 
Caelin Rohiemon

Tutorial 6 Group 2
# Part 1 - Identify a Social Problem



Use APA referencing throughout your document. [https://www.mendeley.com/guides/apa-citation-guide/]

## 1.1 Describe the Social Problem

Artificial Intelligence (AI) is changing the way how firms hire workers, manage and organise work and use technology which has led to an increasingly important social and economic issue. With recent developments in generative AI, especially after the release of Chatgpt, in Novemeber of 2022. AI tools are more visible and widely used across different industries. It is argued that AI in the workplace increase productivity but it also may reduce demand of some type of labour, especially in entry level jobs (Cazzaniga et al ., 2024). 

This becomes increasingly relevant for young workers aged 18-24. As workers in this age group, often have low to nil level of work experience and are very likely to work in temporary, entry level jobs. Therefore, this results in these workers being vulnurable when employers change the skills that they demand or when certain tasks can be fulfilled by AI. Youth Employement is therefore a social problem, because unemployement in early stages of their career/life can negatively affect future earnings,skill development and work experience which is required for future higher skilled jobs (International Labour Organisation, 2024). 

This project examines whether the rise of national AI-Development capcity is linked and associated with worse labour outcomes for young workers compared with the total labour force. We compare data from countries from the year 2021-2025. The aim is whether there is an observable association between AI adoption and youth unemployement relative to overall unemployement. 

Cazzaniga, M., Jaumotte, F., Li, L., Melina, G., Panton, A. J., Pizzinelli, C., Rockall, E. J., Tavares, M. M., & Yao, K. (2024)



*Gen-AI: Artificial intelligence and the future of work*. International Monetary Fund. https://www.imf.org/en/Publications/Staff-Discussion-Notes/Issues/2024/01/14/Gen-AI-Artificial-Intelligence-and-the-Future-of-Work-542379


International Labour Organization. (2024). *Global employment trends for youth 2024*. International Labour Organization. https://www.ilo.org/publications/major-publications/global-employment-trends-youth-2024


Include the following:

- Why is this relevant?

- ...

# Part 2 - Data Sourcing

## 2.1 Load in the data

```{r loading_data and retrieve only data of European countries}

options(timeout = 600)
# Eurostat 
ai_raw <- get_eurostat("isoc_eb_ai", time_format = "num")

# Filter out EU countries, fix the 2 misaligned EU codes and drop aggregates
eu_codes <- ai_raw %>%
  distinct(geo) %>%
  mutate(iso2c = recode(geo, "EL" = "GR", "UK" = "GB")) %>%
  filter(nchar(iso2c) == 2, iso2c != "EA") %>%
  pull(iso2c)

# World Bank: download ONCE, then reuse the saved copy
if (file.exists("wb_raw.rds")) {
  wb_raw <- readRDS("wb_raw.rds")
} else {
  wb_raw <- WDI(
    country   = eu_codes,
    indicator = c(total_unemp = "SL.UEM.TOTL.ZS",
                  youth_unemp = "SL.UEM.1524.ZS"),
    start = 1991, end = 2024, extra = TRUE
  )
  saveRDS(wb_raw, "wb_raw.rds")
}
```

## 2.2 Provide a short summary of the dataset(s)

```{r}
head(ai_raw) 
str(ai_raw)
summary(ai_raw)
names(ai_raw)

head(wb_raw)
str(wb_raw)
summary(wb_raw)
names(wb_raw)

```

## 2.3 Describe the type of variables included

Think of things like:

- Do the variables contain health information or SES information?

- Have they been measured by interviewing individuals or is the data coming from administrative sources?

The AI data comes from Eurostats that shows business wealth and resources . The unemployment data comes from World Bankdata that shows the financial trouble and living standards .the two main variables are  Total unemployment rate and youth unemploymentent. the youth unemployment is used as the sub-population to see how young workers are affected by AI data usage in businesses.
Where: European countries only.
When: Years 2021 to 2025 ( but 2022 not included). This timeline captures the exact period where the post-pandemic labor market met the massive generative AI boom.

*For the sake of this example, I will continue with the assignment...*

# Part 3 - Quantifying

## 3.1 Data cleaning

Say we want to include only larger distances (above 2) in our dataset, we can filter for this.

```{r data_cleaning}

wb_clean <- wb_raw %>%
  filter(year == 2021 | year == 2022 | year == 2023 | year == 2024)

ai_clean <- ai_raw %>%
  filter(indic_is == "E_AI_TANY")

```

Please use a separate 'R block' of code for each type of cleaning. So, e.g. one for missing values, a new one for removing unnecessary variables etc.

## 3.2 Generate necessary variables

Variable 1

```{r gen_var1}

wb_clean <- wb_clean %>%
  mutate(youth_ratio = youth_unemp / total_unemp)

```

Variable 2

```{r gen_var2}
combined <- wb_cleaned %>%
  left_join(ai_clean, by = c("iso2c" = "geo", "year" = "TIME_PERIOD")) %>%
  mutate(ai_group = case_when(
    values < 10 ~ "Low",
    values < 30 ~ "Medium",
    TRUE        ~ "High"
  ))
```

Table with new variables
```{r table}
combined %>%
  group_by(ai_group) %>%
  summarise(
    n = n(),
    mean_total_unemp = mean(total_unemp, na.rm = TRUE),
    mean_youth_ratio = mean(youth_ratio, na.rm = TRUE)
  )
```


## 3.3 Visualize temporal variation
```{r temporal_variation}
yearly <- wb_clean %>%
  group_by(year) %>%
  summarise(
    mean_total_unemp = mean(total_unemp, na.rm = TRUE),
    mean_youth_unemp = mean(youth_unemp, na.rm = TRUE)
  )

ggplot(yearly, aes(x = year)) +
  geom_line(aes(y = mean_total_unemp, color = "Total")) +
  geom_line(aes(y = mean_youth_unemp, color = "Youth (15-24)")) +
  labs(title = "EU unemployment over time: youth vs total",
       x = "Year",
       y = "Unemployment (% of labour force)",
       color = "Group") +
  theme_minimal()
```
ai_group - bins the continuous values into Low, Medium, and High, so we can compare youth unemployment outcomes across different levels of AI adotion.

## 3.4 Visualize spatial variation

```{r visualise map}

ai_group <- combined %>%
   filter(year == 2024, !is.na(ai_group)) %>%
  mutate(ai_group = factor(ai_group, levels = c("Low", "Medium", "High")))

ggplot (combined, aes(x=ai_group, 
                    y= youth_unemp,
                    size= total_unemp,
                    label=country)) +
  geom_point(alpha=0.7, color = "blue") + 
  geom_text_repel(size=3, max.overlaps= 20)+
  scale_size_continuous(name="total unemployment (%)") +
  
  labs(x="AI intensity", 
       y="youth unemployment(%)", 
       title= "AI intensity and unemployment in EU countries") + theme_minimal()

```


## 3.5 Visualize sub-population variation

What is the poverty rate by state?

```{r visualise_subpopulations}
combined %>%
  filter(year == 2024, !is.na(ai_group)) %>%
  mutate(ai_group = factor(ai_group, levels = c("Low", "Medium", "High"))) %>%
  ggplot(aes(x = ai_group, y = youth_unemp)) +
  geom_boxplot(na.rm = TRUE, fill = "antiquewhite", color = "navy", outlier.color = "aquamarine", varwidth = TRUE) +
  labs(title = "Youth unemployment by level of AI adoption (2024)",
       x = "AI adoption group", y = "Youth unemployment (%)") +
  theme_minimal()
```

Here you provide a description of why the plot above is relevant to your specific social problem.




## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis}
ratio_by_year <- combined %>%
  group_by(year) %>%
  summarize(
    mean_youth_ratio = mean(youth_ratio, na.rm = TRUE)
  )
ratio_by_year

#Plotting the youth-to-total unemployment ratio per year 
ggplot(ratio_by_year, aes (x = year, y = mean_youth_ratio)) +
  geom_line() +
  geom_point() +
  geom_vline(xintercept = 2022, linetype = "dashed", color = "red") +
  annotate("text", x = 2022.1, y = 2.8, label = "ChatGPT launch", hjust = 0, color = "red") +
    labs(title = "Youth-to-total unemployment ratio in EU countries (2021-2024)",
         x = "Year",
         y = "Youth-to-total unemployment ratio") +
    theme_minimal()
```
The graph shows the average youth-to-total unemployment ratio across EU countries from 2021 to 2024. The ratio rises steadily after 2022 - the year ChatGPT launched - suggesting that young workers may be increasingly disadvantaged relative to the broader workforce as AI adoption grows. However, this is an association, not proof: other factors such as post-COVID labor market shifts or broader economic conditions could equally explain the trend.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

Use APA referencing throughout your document.

(2026). Europa.eu. https://ec.europa.eu/eurostat/databrowser/view/isoc_eb_ain2/default/table?lang=en


‌