---
title: "Template"
author: "Studentnames and studentnumbers here"
date: "`r Sys.Date()`"
output: pdf_document
---

# Set-up your environment

```{r package_install, include=FALSE}
install.packages("tidyverse")
install.packages("yaml")
install.packages("rmarkdown")
install.packages("WDI")
install.packages("eurostat")
```

```{r packages}
require(tidyverse)
require(rmarkdown)
require(yaml)
require(WDI)
require(eurostat)
```

# Title Page

<<<<<<< Updated upstream
LE Vi Trang Raghoebar Monisha Isabel Lust Ermias Gebremichael Pritish Dewan Pritish Dewan Noel Tripathi Caelin Rohiemon
=======
LE Vi Trang Raghoebar Monisha Isabel Lust Ermias Gebremichael Pritish Dewan Noel Tripathi Caelin Rohiemon
>>>>>>> Stashed changes


Tutorial 6 Group 2

Simon Loop

# Part 1 - Identify a Social Problem

Use APA referencing throughout your document. [Here's a link to some explanation.](https://www.mendeley.com/guides/apa-citation-guide/)

## 1.1 Describe the Social Problem

Include the following:

- Why is this relevant?

- ...

# Part 2 - Data Sourcing

## 2.1 Load in the data

```{r loading_data and retrieve only data of European countries}

options(timeout = 600)
# Eurostat 
ai_raw <- get_eurostat("isoc_eb_ai", time_format = "num")

# Filter out EU countries, fix the 2 misaligned EU codes and drop aggregates
eu_codes <- ai_raw %>%
  distinct(geo) %>%
  mutate(iso2c = recode(geo, "EL" = "GR", "UK" = "GB")) %>%
  filter(nchar(iso2c) == 2, iso2c != "EA") %>%
  pull(iso2c)

# World Bank: download ONCE, then reuse the saved copy
if (file.exists("wb_raw.rds")) {
  wb_raw <- readRDS("wb_raw.rds")
} else {
  wb_raw <- WDI(
    country   = eu_codes,
    indicator = c(total_unemp = "SL.UEM.TOTL.ZS",
                  youth_unemp = "SL.UEM.1524.ZS"),
    start = 1991, end = 2024, extra = TRUE
  )
  saveRDS(wb_raw, "wb_raw.rds")
}
```

## 2.2 Provide a short summary of the dataset(s)

```{r}
head(ai_raw) 
str(ai_raw)
summary(ai_raw)
names(ai_raw)

head(wb_raw)
str(wb_raw)
summary(wb_raw)
names(wb_raw)

```

## 2.3 Describe the type of variables included

Think of things like:

- Do the variables contain health information or SES information?

<<<<<<< Updated upstream
- Have they been measured by interviewing individuals or is the data coming from administrative sources?

The AI data comes from Eurostats that shows business wealth and resources . The unemployment data comes from World Bankdata that shows the financial trouble and living standards .the two main variables are  Total unemployment rate and youth unemploymentent. the youth unemployment is used as the sub-population to see how young workers are affected by AI data usage in businesses.
Where: European countries only.
When: Years 2021 to 2025 ( but 2022 not included). This timeline captures the exact period where the post-pandemic labor market met the massive generative AI boom.
<<<<<<< Updated upstream
=======

=======

>>>>>>> Stashed changes
 
 
*For the sake of this example, I will continue with the assignment...*
>>>>>>> Stashed changes

# Part 3 - Quantifying

## 3.1 Data cleaning

Say we want to include only larger distances (above 2) in our dataset, we can filter for this.

```{r data_cleaning}

wb_cleaned <- wb_raw %>%
  filter(year == 2021 | year == 2022 | year == 2023 | year == 2024)

ai_clean <- ai_raw %>%
  filter(indic_is == "E_AI_TANY")
  
<<<<<<< Updated upstream
=======




>>>>>>> Stashed changes

```

Please use a separate 'R block' of code for each type of cleaning. So, e.g. one for missing values, a new one for removing unnecessary variables etc.

## 3.2 Generate necessary variables

Variable 1

```{r gen_var1}

wb_cleaned <- wb_cleaned %>%
  mutate(youth_ratio = youth_unemp / total_unemp)

```

Variable 2

```{r gen_var2}

<<<<<<< Updated upstream
combined <- wb_clean %>%
=======
combined <- wb_cleaned %>%
>>>>>>> Stashed changes
  left_join(ai_clean, by = c("iso2c" = "geo", "year" = "TIME_PERIOD")) %>%
mutate(ai_group = case_when(
  values < 10 ~ "Low",
  values < 30 ~ "Medium",
  TRUE ~ "High"
))
<<<<<<< HEAD
names(wb_clean)

```
<<<<<<< Updated upstream
=======
=======
names(wb_cleaned)

```
>>>>>>> Stashed changes

>>>>>>> 803be7339e7448133311b0708fd03799f78c06d4

Table with new variables

```{r}
combined %>%
  group_by(ai_group) %>%
  summarise(
    n = n(),
    mean_total_unemp = mean(total_unemp, na.rm = TRUE),
    mean_youth_ratio = mean(youth_ratio, na.rm = TRUE)
  )
```



## 3.3 Visualize temporal variation
<<<<<<< Updated upstream
=======



>>>>>>> Stashed changes
```{r}
ggplot(wb_cleaned, aes(x= year)) + 
         geom_line(aes(y=mean_total), color = "red") +
          geom_line(aes(y=mean_youth), color = "blue")
```
ai_group - bins the continuous values into Low, Medium, and High, so we can compare youth unemployment outcomes across different levels of AI adotion.


## 3.4 Visualize spatial variation

```{r visualise map}

<<<<<<< Updated upstream
ai_group <- combined %>%
   filter(year == 2024, !is.na(ai_group)) %>%
  mutate(ai_group = factor(ai_group, levels = c("Low", "Medium", "High")))

ggplot (combined, aes(x=ai_group, 
                    y= youth_unemp,
                    size= total_unemp,
                    label=country)) +
  geom_point(alpha=0.7, color = "blue") + 
  geom_text_repel(size=3, max.overlaps= 20)+
  scale_size_continuous(name="total unemployment (%)") +
  
  labs(x="AI intensity", 
       y="youth unemployment(%)", 
       title= "AI intensity and unemployment in EU countries") + theme_minimal()

=======
>>>>>>> Stashed changes
```


## 3.5 Visualize sub-population variation

What is the poverty rate by state?

```{r visualise_subpopulations}
<<<<<<< Updated upstream
combined %>%
  filter(year == 2024, !is.na(ai_group)) %>%
  mutate(ai_group = factor(ai_group, levels = c("Low", "Medium", "High"))) %>%
  ggplot(aes(x = ai_group, y = youth_unemp)) +
  geom_boxplot(na.rm = TRUE, fill = "antiquewhite", color = "navy", outlier.color = "aquamarine", varwidth = TRUE) +
  labs(title = "Youth unemployment by level of AI adoption (2024)",
       x = "AI adoption group", y = "Youth unemployment (%)") +
  theme_minimal()

=======
dataset$inmetro <- dataset$inmetro %>% as.factor()
# Boxplot of poverty rate by state using the 'midwest' dataset
ggplot(dataset, aes(x = inmetro, y = percadultpoverty)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Poverty Rates by Metropolitan status (Midwest counties)",
    x = "Metropolitan Area",
    y = "Poverty Rate of Adults (%)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right"
  )
>>>>>>> Stashed changes
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis}
ratio_by_year <- combined %>%
  group_by(year) %>%
  summarize(
    mean_youth_ratio = mean(youth_ratio, na.rm = TRUE)
  )
ratio_by_year

#Plotting the youth-to-total unemployment ratio per year 
ggplot(ratio_by_year, aes (x = year, y = mean_youth_ratio)) +
  geom_line() +
  geom_point() +
  geom_vline(xintercept = 2022, linetype = "dashed", color = "red") +
  annotate("text", x = 2022.1, y = 2.8, label = "ChatGPT launch", hjust = 0, color = "red") +
    labs(title = "Youth-to-total unemployment ratio in EU countries (2021-2024)",
         x = "Year",
         y = "Youth-to-total unemployment ratio") +
    theme_minimal()
```
The graph shows the average youth-to-total unemployment ratio across EU countries from 2021 to 2024. The ratio rises steadily after 2022 - the year ChatGPT launched - suggesting that young workers may be increasingly disadvantaged relative to the broader workforce as AI adoption grows. However, this is an association, not proof: other factors such as post-COVID labor market shifts or broader economic conditions could equally explain the trend.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

Use APA referencing throughout your document.
